---
title: "ESM 244 Lab 1"
author: "Kiera Matiska"
date: "1/4/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)
# install.packages('tmap')
# update.packages(ask = FALSE)
```

### Read in the data

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"),
                     show_col_types = FALSE)
```

### Part 1: wrangling and ggplot review

**Example 1:** Find counts of observation by `legal_status` & wrangle a bit

```{r}
# method 1: group_by() %>% summarize()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

# method 2: different way plus a few new functions
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% 
  slice_max(tree_count, n = 5) %>% 
  arrange(desc(tree_count)) # or arrange(-tree_count)
```

Make a graph of the top 5 from above:

```{r}
ggplot(data = top_5_status,
       aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) +
  geom_col(fill = "darkgreen") +
  labs(x = "Legal Status",
       y = "Tree Count") +
  coord_flip() +
  theme_minimal()
```

**Example 2:** Only going to keep observation where legal status is "Permitted Site" and caretaker is "MTA", and store as `permitted_data_df`

```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker)
permitted_data_df <- sf_trees %>% 
  filter(legal_status == "Permitted Site", caretaker == "MTA") # can also use & in replace of the comma
```

**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `latituse`, `longitude`, and store as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <-sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)

# make a little graph of locations
ggplot(data = blackwood_acacia_df, 
       aes(x = lon, y = lat)) +
  geom_point(color = "darkgreen")
```

**Example 4:** Use `tidyr::separate()`

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific", "spp_common"), sep = " :: ")
```

**Example 5:** use `tidyr::unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite("id_status", tree_id, legal_status, sep = "_COOL_")
```

